Estatus social subjetivo, posición objetiva y percepción de meritocracia: una aproximación empírica al caso de Chile

### 2023-01-12 paper-blind.Rmd
ajustes de introduccion

---
 title: "Status, Class, Subjective status and meritocracy perception"
 # author: Julio Iturra ^[Bremen International Graduate School of Social Sciences (jiiturra@uni-bremen.de)]
 bibliography: "input/bib/ess-merit.bib"
 csl: "input/bib/apa6.csl"
 css: "input/css/custom.css"
 output:
     bookdown::pdf_document2:
         template: default
         toc: false
         number_sections: false
     bookdown::html_document2:
          number_sections: true
     bookdown::word_document2:
          toc: true
 geometry: margin=1in
 header-includes:
   # - \usepackage[spanish,es-tabla,es-nodecimaldot]{babel} # para decimales con commas en vez de punto
   - \usepackage{times}           # Times New Roman
   - \usepackage{caption}
   - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
   - \usepackage{graphicx}
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
 linestretch: '1.5'
 fontsize: 12pt
 link-citations: yes
 linkcolor: blue
 editor_options:
   chunk_output_type: inline
 abstract: "What is the role of status, class and subjective status on perceived meritocracy? To what extent can subjective status affect the objective experiences given by income, education and social class on meritocracy perception? To answer these questions, we use nationally representative data from the 2018 wave of the Longitudinal Social Survey of Chile - ELSOC (N = 2,652). Perceived meritocracy constitutes the descriptive views about how effort and talent are rewarded in society. Contrary to our expectations, income and social class do not influence perceived meritocracy. However, in line with the hypothesis, individuals with higher educational attainment hold critical views about meritocracy, and those who allocate themselves close to the top groups in society perceive higher rewards regarding effort and talent. We also found that subjective status increases the meritocracy perception of those groups on top of the social hierarchy concerning income, education and social class."  
---

```{r render,include=FALSE, eval=FALSE}
 rmarkdown::render("paper-blind.Rmd")
```

@@

@@

In recent decades, increasing economic inequality [@piketty_capital_2014] and stagnant social mobility [@oecd_broken_2018] have prompted social scientists to examine how societies perceive the functioning of meritocracy - the idea that effort and talent are rewarded - and its role in justifying economic inequality [@mijs_meritocracy_2020]. Meritocracy has been linked to justifying social inequalities, as individual achievements and class position are often attributed to individual agency rather than structural constraints [@frei_fruto_2020]. Studies have suggested that individuals' views of the labor market shape their perception of how economic resources are distributed [@Svallfors2006]. Belonging to non-manual and professional social classes is negatively associated with perceived meritocracy, especially in societies with high levels of economic inequality and low social mobility [@Canales2015lse].

Social stratification can be studied objectively, through the expression of social inequality in economic and cultural resources [@Kerbo2003], or subjectively, through individual experiences in the class structure [@Kluegel-Smith1986]. Class is often linked to occupation, labor market position, and levels of authority, while status refers to perceived superior, equal, or inferior relationships among individuals [@Chan2007]. Although there are scales of subjective social class [@Elbert2018], subjective social status has been more widely studied [@zhou_trajectory_2021]. Subjective social status refers to an individual's or group's perception of their standing in the social hierarchy compared to a reference group, such as society [@Evans2004] or their local community [@condon_economic_2020], and can influence perceptions of meritocracy [@Castillo2018]. People tend to compare themselves to similar groups to reduce cognitive dissonance and better estimate their societal position [@Merton1968;@festinger1954theory]. High-status individuals tend to underestimate their position while low-status individuals overestimate it [@Cruces2011]. The characteristics of an individual's reference groups can shape their subjective estimations and their image of society is influenced by the homophily of their close networks [@Evans2017]. Higher subjective social status is associated with higher perceived meritocracy and can also positively moderate the influence of class and status [@hvidberg_social_2020].

This study aims to understand the influence of objective class and status on meritocracy perception, as well as the role of subjective social status and its interaction with social class, education, and income.

## Meritocracy perception

Merit is understood as the combination of effort and ability [@Sen2000]. Thus, the meritocratic ideal is focused on the individual, rather than structural factors in getting ahead in life [@mccall_exposure_2017].  In this regard, the measurement of meritocracy-related constructs has not been clear in distinguishing between descriptive and normative dimensions [@Castilloetal2019]. For instance, it is often to notice undifferentiated use of *Meritocratic Beliefs* [@mijs_how_2021a; @bernardo_meritocracy_2021] and *Perceived Meritocracy* [@Xian2017]. The lack of distinction can be problematic, especially when scrutinizing explanations on subjective views about meritocracy.

The concept of  _perception_ is understood as the subjective description of an object [@Aalberg2003]. In this regard, the descriptive (*how is*) dimension does not consider the normative evaluation of the object (*should be*). On the other hand, the normative dimension can be understood as preferences, influenced by previous beliefs and information affecting the evaluation of an object [@druckman_preference_2000]. In this regard, @Garcia2001 suggests perceived meritocracy is the degree to which individuals consider their society complies with the meritocratic principles. In other words, to what extent effort and talent are rewarded, indicating "how meritocracy is" rather than "how it should be".

## Status, Class and meritocracy perception

### Status

According to the rational self-interest approach [@Meltzer81], individuals with higher incomes perceive higher meritocracy because they accumulated economic resources through effort and talent. High status is also associated with lower inequality perception [@Evans2017] and higher justification of income inequality [@Trump2017].

There is evidence of a positive relationship between income and perceptions of meritocracy regarding the importance of individual merit in achieving success [@Reynolds2014; @Xian2017] and the extent to which intelligence and skills are rewarded [@Castilloetal2019; @Duru-bellat2012]. However, there is mixed evidence on how lower-income individuals rationalize inequality and meritocracy. Some research supports the "activated conflict theory," which suggests that the poor in unequal localities hold lower beliefs in meritocracy due to social comparisons that increase awareness of social status [@Newman2015]. Other research, based on the "relative power theory," challenges this finding, arguing that the rich hold more power due to income and wealth concentration, allowing them to disseminate legitimizing ideas about economic inequality, and the poor are less critical of dominant beliefs like meritocracy as a psychological mechanism for coping with deprivation [@Solt2016]. However, other research shows that low-income individuals in highly unequal contexts have firmer meritocracy beliefs, but the differences tend to fade as inequality declines [@morris_paradox_2022].

> $H_{\text{1a}}$: Income is positively correlated with perceived meritocracy.

As income reflects an individual's economic resources, studies on social stratification have raised the importance of educational attainment on status achievement [@Goldthorpe2003]. In this line, research on meritocracy perception have discussed two theoretical approaches. On the one hand, the *reproductionist* approach [@Bourdieu2009; @bernstein_class_2003] supports the *socialisation* hypothesis, suggesting school socialisation operates as a normative reinforcement of meritocratic ideals, then educated individuals are more willing to support meritocracy as a legitimising mechanism [@Lampert2013]. On the other hand, the *instruction* hypothesis suggests perceived meritocracy tends to weaken in individuals with higher educational attainment because they would be more aware of the influence of structural factors on status achievement [@Duru-bellat2012] considering experiences of educational mobility [@Canales2015lse] and higher chances to engage with complex political ideas [@bubak_perceptions_2019].

On the one hand, evidence in favour of the socialisation hypothesis has pointed out education is associated with higher importance of individual factors in getting ahead in life [@Xian2017; @Reynolds2014; @mijs_paradox_2021], but also as a determinant of wages [@Kunovich2007]. On the other hand, evidence in favour of the instruction hypothesis shows that individuals more educated individuals disagree that people are rewarded for their effort and talent [@Duru-bellat2012; @Castilloetal2019]. Additionally, the instructional hypothesis can be complemented by the fact that greater awareness of inequality may experience feelings of injustice and relative deprivation since individuals with the same educational attainment have access to unequal rewards [@Adams1965; @Smith-eta2020_relative-deprivation]. In this sense, it has been shown higher educational attainment is associated with higher perceived inequality [@Castillo2012a_justice]. Thus, this evidence lead us to suggest:

> $H_{\text{1b}}$: education is negatively associated with perceived meritocracy.

### Class

The relationship between structural position and meritocracy has mainly focused on income and educational attainment, setting social class aside from the discussion. However, occupational class not only represent the synthesis of educational achievement and income levels [@Ganzeboom1992] but also represents a symbol of power and authority in contemporary societies [@Chan2007]. Furthermore, studies on social mobility have shown the relevance of studying inequality through class schemes using occupation, autonomy and authority in the workspace as meaningful distinctions to categorise individuals [@Erikson1992], making them relatively homogeneous groups   [@Connelly2016]. However, it has been argued that common characteristics do not preclude class heterogeneity in terms of cultural consumption behaviour or political preferences [@Chan2004].

The literature on distributive preferences has suggested social class reflects individuals' experience in the labour market regarding job opportunities and wages [@Svallfors2006]. In a similar vein, @Kulin2013 demonstrated lower classes being exposed to higher labour market risks, such as the likelihood of job loss and lower wages, led individuals in material deprivation and risk exposure, to demand more redistribution [@rehm_risks_2009].

The literature has paid little attention to social class on meritocracy perception. For example, @Smith2012 found that individuals belonging to the petty bourgeoisie perceive higher levels of meritocracy compared to wage earners. These findings are supported by @Sandoval2017, which found that skilled and service workers attribute more importance to effort in achieving success compared to unskilled workers. However, @Canales2015lse found that upwardly mobile individuals in the service class perceived lower levels of meritocracy

Despite high income inequality, social mobility is still high in Chile, particularly for individuals from lower-class origins moving to non-manual and service-class destinations [@Torche2005a]. However, wages have remained stagnant and economic inequality has remained relatively stable [@EspinozaNunez2014]. This can lead to negative perceptions of meritocracy due to frustrating income expectations or experiences of discrimination in the labor market. For example, individuals with a lower-class background who have experienced upward mobility [@barozet_clase_2011] often report feeling frustrated by discrimination based on their class background [@cea_discriminacion_2020]. Upward mobility may not always result in satisfactory outcomes, which can further negative attitudes toward labor market constraints. Overall, we can hypothesis that:

> $H_{\text{1c}}$: Working in non-manual and service occupations is negatively associated with perceived meritocracy.

## Subjective social status and meritocracy perception

The literature on subjective meritocracy has paid more attention to the structural position of individuals as an explanatory factor for perceptions of meritocracy. However, how individuals experience social inequality is affected by the status of the individuals' or groups [@castillo_perception_2022]. For example, perceived economic inequality in everyday life diminishes justification of inequality and increases demand for redistribution [@garcia-castro_perceiving_2020]. Nevertheless, it has been demonstrated meritocracy primes reinforce negative evaluations of economically disadvantaged groups and strengthens internal attribution [@madeira_primes_2019].

@@

Social perception of meritocracy may be influenced by an individual's reference group and their perceived social standing within that group. The reference group and reality blend (R&R) hypothesis suggests that an individual's cognitive biases may be shaped by an availability heuristic, reducing cognitive dissonance in situations with limited information [@KelleyEvans1995; @Evans2004]. Close reference groups can shape an individual's subjective estimates of their social standing, and their imagined ideas about society may be influenced by the homophily within their social networks. Research has shown that individuals with networks consisting of people with high-status occupations have a higher subjective social status in Chinese society, where social capital is important [@kim_social_2021]. Membership in organizations or participation in hobbies may also shape cognitive patterns of homophily and influence an individual's perceived social standing [@son_cognitive_2021].

The link of subjective status with attitudes toward inequality has been little studied. For example, @Vargas-Salfate2018 found that higher subjective status is associated with higher system justification. Similarly, @Schneider2015 evidence a positive association of subjective status with the perceived justice income inequality and higher internal poverty attributions. In addition, @Castillo2018 found that higher subjective status is related to higher perceived meritocracy. Based on this evidence, our main prediction is:

> $H_{\text{2}}$: subjective social status is positively associated with perceived meritocracy

## Objective and subjective position on perceived meritocracy

@@

Higher-income individuals tend to have higher inconsistencies in perceived status [@Castillo2013]. For example, income, education, and occupational prestige have a positive non-linear relationship with subjective social status, with higher status individuals showing more variation in subjective status [@Andersson2018a]. Inconsistencies between income level and educational attainment have been linked to political preferences [@Kreckel1993, @Lenski1967], and discrepancies between class identity and social class may also influence political preferences [@Sosnaud2013], but these results are not consistently supported by socio-demographic controls.

Studies on subjective social status as a moderating factor have primarily focused on its relationship to other subjective variables, rather than how it interacts with income or education. One study using data from the International Social Survey Programme (ISSP) found that higher perceptions of inequality lead to higher preferences for redistribution and lower perceptions of meritocracy, and that the relationship between inequality and meritocracy perception becomes weaker for those who perceive themselves as having a higher societal position [@Fatke2018]. Another study in European countries showed that economic inequality does not influence preferences for redistribution when controlled by perceived inequality, but is positively moderated by subjective status [@Bobzien2019]. A study on bachelor students with high subjective social status found that those who perceive themselves as having a higher status and greater meritocracy tend to justify higher levels of economic inequality [@bernardo_meritocracy_2021].

@@

Studies in Chile have found that three factors interact to shape perceptions of social status: socioeconomic status, economic and social capital, and moral perceptions of individual achievement [@macclure_naming_2020]. Individuals who identify as lower status tend to have low incomes, less education, and live in disadvantaged areas. However, some individuals who belong to middle or high socioeconomic strata may still identify as lower status due to living in affluent neighborhoods or having higher educational credentials, which symbolize higher social and cultural capital [@Mclure-etal2019, p. 18]. Those who identify as high status tend to consistently have higher socioeconomic status, with educational meritocracy often serving as a criterion for their social position [@Mclure-etal2019,p.19].

Subjective social status can reflect an individual's position in the social structure as well as their experience of economic inequality through comparisons with nearby reference groups. Therefore, we predict that differences in subjective status can lead to differences in perceived meritocracy within socio-economic groups. Thus, we predict that:

@@

The data used correspond to the third wave of the Longitudinal Social Study of Chile [@coes_estudio_2018], a survey developed to analyse the evolution of conflict and social cohesion in Chilean society over time. The sampling is probabilistic, stratified, clustered and multistage, with a total of 3748 participants aged between 18 and 75 years. An analytical sample of 2652 cases with complete information is used for the analyses.

@@

The objective position is measured through three variables. First, _Household Income Decile_ is calculated using monthly per capita household income and categorised into deciles.  Second, _Educational level_ corresponds to the five categories for educational attainment following the International Standard Classification of Education [@unesco_international_2012]. Finally, we use an adapted five-category version of the EGP class scheme [@Connelly2016], a commonly used measurement in social stratification research [@solis_class_2019].  The Stata package `isko` was used for the computation of the class scheme [@hendrickx_isko_2002].

@@

@@

  df_desc<- elsoc_w03 %>%
    dplyr::select(merit,meffort,mtalent,
           inc10h.imp,edcine2,egpisko05,ess,lngap_perc,
           sexo,edad,ppolcat) %>% 
    dplyr::select(-meffort,-mtalent,
                  # -sexo,-edad,
                  # -ppolcat,-lngap_perc
                  )
  varlabs <- sjlabelled::get_label(df_desc) #recuperar labels
  df_desc <- na.omit(df_desc)
  set_label(df_desc) <- varlabs  
 st_options(lang = "en",
            footnote = paste0("Source: Authors calculation based on ELSOC 2018 (N = ",
            dim(df_desc)[[1]],")"),
            bootstrap.css = F,
            # custom.css = "input/css/dfsummary.css"
            )
 df<- dfSummary(df_desc,
                plain.ascii = FALSE,
                style = "grid",
                tmp.img.dir = "/tmp",
                graph.magnif = TRUE,
                headings = F,  # encabezado
                varnumbers = F, # num variable
                labels.col = T, # etiquetas
                na.col = F,    # missing
                graph.col = T, # plot
                valid.col = T,max.string.width = 50,
                max.distinct.values = 8 # n valido 
                # col.widths = c(1000,10,10,10)
                )
 df$Variable <- NULL # delete variable column
 df$Valid <- NULL
 summarytools::view(df,file = "output/tables/desc01.html") # Ver tabla en un archivo HTML
 webshot(url ="output/tables/desc01.html",
         file ="output/tables/desc01.png",
         vwidth = 700,vheight = 600) # Remover archivo HTML
```

@@

Considering the dependent variable as a continuous, stepwise ordinary least square regression models are used (OLS). First, models are estimated for each of the independent variables. Second, a series of models are estimated that consider the interaction term between objective position and subjective status. Interactions effects are presented using conditional linear predicted values <!-- and linear predictions -->.

@@

Figure \@ref(fig:corr-mat) shows that perceived meritocracy has a weak positive correlation with income decile, but weak negative correlations with education and social class. The correlation between perceived meritocracy and subjective social status is positive, with a slightly stronger association for the indicator for effort (r = 0.08) compared to intelligence (r = 0.06).

@@

# Dicussion and conclusion

>>>>

---
 title: "Subjective status, Social Class and meritocracy perception"
 # author: Julio Iturra ^[Bremen International Graduate School of Social Sciences (jiiturra@uni-bremen.de)]
 bibliography: "input/bib/ess-merit.bib"
 csl: "input/bib/apa6.csl"
 css: "input/css/custom.css"
 output:
     bookdown::html_document2:
          number_sections: false
     bookdown::pdf_document2:
         template: default
         toc: false
         number_sections: false
     bookdown::word_document2:
          toc: true
 geometry: margin=1in
 header-includes:
   # - \usepackage[spanish,es-tabla,es-nodecimaldot]{babel} # para decimales con commas en vez de punto
   - \usepackage{times}           # Times New Roman
   - \usepackage{caption}
   - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
   - \usepackage{graphicx}
   - \usepackage{float}
   - \usepackage{booktabs}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
 linestretch: '1.5'
 fontsize: 12pt
 link-citations: yes
 linkcolor: blue
 editor_options:
   chunk_output_type: inline
 abstract: "Perceived meritocracy is understood as how societies describe the functioning of meritocracy, the ideal that promotes the distribution of resources in society based on individual effort and ability. Said this, are subjective and objective positions associated with perceived meritocracy? To what extent is objective position's influence on meritocracy perception moderated by subjective status? To answer these questions, we use nationally representative data from the 2018 wave of the Longitudinal Social Survey of Chile (N = 2,652). Regression analysis show that subjective social status has a robust positive association with perceived meritocracy, in line with the hypothesis. However, both positive and negative associations of income and occupational social class on perceived meritocracy were not confirmed. However, the hypothesis for education was confirmed, where higher educational attainment is associated with lower perceived meritocracy. Finally, we found that subjective status increases perceived meritocracy for those with higher incomes, tertiary education, and higher occupational social class."  
---

```{r render,include=FALSE, eval=FALSE}
 rmarkdown::render("paper-blind.Rmd", output_format = "bookdown::pdf_document2")
 rmarkdown::render("paper-blind.Rmd", output_format = "bookdown::html_document2")
 rmarkdown::render("paper-blind.Rmd", output_format = "bookdown::word_document2")   
```

@@

* Word count: 6908

@@

Meritocracy is an ideal that promotes the distribution of resources in a society based on individual effort and ability [@Young1961]. In modern societies, merit is commonly embraced as a fair way of achieving status, in opposition to drawing on connections or inheritance [@Hadjar2008]. The increasing levels of economic inequality [@piketty_capital_2014] and the stagnation of social mobility [@oecd_broken_2018] has increased the attention of social scientist on the subjective aspects of meritocracy, mainly how societies perceive the functioning of meritocracy and its role in justifying economic inequality [@mijs_meritocracy_2020]. The literature on subjective meritocracy has pointed out the distinction between the descriptive dimension represented by perception (how it is) and preference as the normative dimension (how it should be) [@Castilloetal2019]. The role of social position has been claimed as a key factor in perceived meritocracy by social-psychological studies focusing on subjective social status [@bernardo_meritocracy_2021;@madeira_primes_2019]. In the sociological literature, attention has been paid to income and education as objective status characteristics [@Duru-bellat2012;@Castilloetal2019], but also little consideration of authority and control over labour represented by occupational social class [@Smith2012]. Both disciplinary approaches have been developing parallel research agendas on perceived meritocracy without an extensive dialogue.

Studies have shown the positive influence of economic self-interest [@morris_paradox_2022] but also the negative association between education [@Duru-bellat2012] and occupational social class [@Canales2015lse] with meritocracy perception. In other words, income increases the perception of meritocracy, but education and social class decrease it. Regardless of the materialist explanations of perceived meritocracy [@Goldthorpe2003], subjective social status and its anchoring in social comparison processes is another less-discussed social-psychological factor. The perceived standing in the social hierarchy of an individual or group depends on reference groups, such as top-bottom groups in society [@Evans2004] or their neighbours and local community [@condon_economic_2020], and it has been demonstrated that it can positively influence perceived meritocracy [@Castilloetal2019]. Furthermore, the reference group and reality blend (R&R) hypothesis [@Evans2004] has suggested that status perception and its discrepancies with objective position can be explained by the characteristics of reference groups involved in the comparison, arguing that status perception can lead to different outcomes than objective measures. Thus, what is the role of subjective social status on meritocracy perception? Furthermore, to what extent can the influence of objective position on perceived meritocracy be moderated by subjective status? We argue that subjective social status represents an independent process able to explain why some people perceive lower (or higher) meritocracy even when they hold (dis)advantageous positions in terms of income, educational attainment or occupational social class.

The present research aims to scrutinise the role of both subjective and objective social positions on perceived meritocracy. Regarding subjective social status, we claim that social comparison processes and reference groups play a crucial role in how individuals perceive meritocracy as an independent force. On the side of material factors, we examine the role of income and education as different status characteristics, but we also discussed the less addressed role of occupational social class. Finally, as an attempt to test the reference group and reality blend hypothesis, we tested the hypothesised positive moderating effect of subjective social status on the influence of objective status and social class.

## Perception of meritocracy

<!-- Merit is understood as the combination of effort and ability [@Sen2000]. Thus, the meritocratic ideal is focused on the individual, rather than structural factors in getting ahead in life [@mccall_exposure_2017].  In this regard, the measurement of meritocracy-related constructs has not been clear in distinguishing between descriptive and normative dimensions [@Castilloetal2019]. For instance, it is often to notice undifferentiated use of *Meritocratic Beliefs* [@mijs_how_2021a; @bernardo_meritocracy_2021] and *Perceived Meritocracy* [@Xian2017]. The lack of distinction can be problematic, especially when scrutinizing explanations on subjective views about meritocracy.   -->

<!-- The concept of  _perception_ is understood as the subjective description of an object [@Aalberg2003]. In this regard, the descriptive (*how is*) dimension does not consider the normative evaluation of the object (*should be*). On the other hand, the normative dimension can be understood as preferences, influenced by previous beliefs and information affecting the evaluation of an object [@druckman_preference_2000]. In this regard, @Garcia2001 suggests perceived meritocracy is the degree to which individuals consider their society complies with the meritocratic principles. In other words, to what extent effort and talent are rewarded, indicating "how meritocracy is" rather than "how it should be".  -->

Merit is understood as the combination of individual effort and ability [@Sen2000]. Moreover, the meritocratic ideal assumes that status achievement is not constrained by how opportunities are structured in society [@mccall_exposure_2017]. Both constitute the substantive core of how meritocracy can be comprehended. However, studies on meritocracy have made undifferentiated use of concepts such as *Meritocratic Beliefs* [@mijs_how_2021a; @bernardo_meritocracy_2021] or *Perceived Meritocracy* [@Duru-bellat2012]. This lack of conceptual and empirical accuracy is often noticed in studies addressing meritocracy's subjective aspect without distinguishing between descriptive and normative dimensions [@Castilloetal2019]. The research on subjective meritocracy has focused on structural inequality as an explanatory factor, mainly analysing the role of social status, reporting mixed results given by different measurement strategies of the concept.

The concept of  _perception_ is understood as the subjective description of an object [@Janmaat2013]. In this regard, the descriptive dimension (*how is*) does not consider the normative evaluation of the object (*should be*) understood as preferences [@Janmaat2013]. In this regard, @Garcia2001 suggests perceived meritocracy is the degree to which individuals consider their society complies with the meritocratic principles. In other words, to what extent effort and talent are rewarded, indicating "how meritocracy is" rather than "how it should be". Previous evidence suggests that preferences, compared to perceptions, are more stable across status groups [@Castilloetal2019,p.10]. Thus, we expect that subjective and objective social positions can explain the differences in meritocracy perception.

## Subjective social status

The literature on subjective meritocracy has paid more attention to the structural position of individuals as an explanatory factor for perceptions of meritocracy. How individuals experience social inequality is affected by the status of the individuals' or groups [@castillo_perception_2022]. For example, perceived economic inequality in everyday life diminishes justification of inequality and increases demand for redistribution [@garcia-castro_perceiving_2020]. Nevertheless, it has been demonstrated that meritocracy primes reinforce negative evaluations of economically disadvantaged groups and strengthens internal attribution [@madeira_primes_2019].

@@

Social perception of meritocracy may be influenced by an individual's reference group and their perceived social standing within that group. The reference group and reality blend (R&R) hypothesis suggests that an availability heuristic may shape an individual's cognitive biases, reducing cognitive dissonance in situations with limited information [@KelleyEvans1995; @Evans2004]. Close reference groups can shape an individual's subjective estimates of their social standing, and the homophily within their social networks may influence their imagined ideas about society. For example, research has shown that individuals with networks of people with high-status occupations have a higher subjective social status in Chinese society, where social capital is important [@kim_social_2021]. Membership in organisations or participation in hobbies may also shape cognitive patterns of homophily and influence an individual's perceived social standing [@son_cognitive_2021].

The link of subjective status with attitudes toward inequality has been little studied. For example, @Vargas-Salfate2018 found that higher subjective status is associated with higher system justification. Similarly, @Schneider2015 evidence a positive association of subjective status with the perceived justice income inequality and higher internal poverty attributions. In addition, @Castilloetal2019 found that higher subjective status is related to higher perceived meritocracy. Based on this evidence, our main prediction is:

> $H_{\text{1}}$: subjective social status is positively associated with perceived meritocracy

## Income

According to the rational self-interest approach [@Meltzer81], higher-income individuals perceive higher meritocracy because they have accumulated economic resources through effort and talent. High status is also associated with lower inequality perception [@Evans2017] and higher justification of income inequality [@Trump2017].

There is evidence of a positive relationship between income and perceptions of meritocracy regarding the importance of individual merit in achieving success [@Reynolds2014; @Xian2017] and the extent to which intelligence and skills are rewarded [@Castilloetal2019; @Duru-bellat2012]. However, there is mixed evidence on how lower-income individuals rationalise inequality and meritocracy. Some research supports the "activated conflict theory," which suggests that the poor in unequal localities hold lower beliefs in meritocracy due to social comparisons that increase awareness of social status [@Newman2015]. Other research, based on the "relative power theory," challenges this finding, arguing that the rich hold more power due to income and wealth concentration, allowing them to disseminate legitimising ideas about economic inequality, and the poor are less critical of dominant beliefs like meritocracy as a psychological mechanism for coping with deprivation [@Solt2016]. However, other research shows that low-income individuals in highly unequal contexts have firmer meritocracy beliefs, but the differences tend to fade as inequality declines [@morris_paradox_2022].

> $H_{\text{2a}}$: Income is positively correlated with perceived meritocracy.

## Education

As income reflects an individual's economic resources, studies on social stratification have raised the importance of educational attainment on status achievement [@Goldthorpe2003]. In this line, research on meritocracy perception has discussed two theoretical approaches. On the one hand, the *reproductionist* approach [@Bourdieu2009; @bernstein_class_2003] supports the *socialisation* hypothesis, suggesting school socialisation operates as a normative reinforcement of meritocratic ideals, then educated individuals are more willing to support meritocracy as a legitimising mechanism [@Lampert2013]. On the other hand, the *instruction* hypothesis suggests perceived meritocracy tends to weaken in individuals with higher educational attainment because they would be more aware of the influence of structural factors on status achievement [@Duru-bellat2012] considering experiences of educational mobility [@Canales2015lse] and higher chances to engage with complex political ideas [@bubak_perceptions_2019].

On the one hand, evidence in favour of the socialisation hypothesis has pointed out that education is associated with higher importance of individual factors in getting ahead in life [@Xian2017; @Reynolds2014; @mijs_paradox_2021], but also as a determinant of wages [@Kunovich2007]. On the other hand, evidence in favour of the instruction hypothesis shows that more educated individuals disagree that people are rewarded for their effort and talent [@Duru-bellat2012; @Castilloetal2019]. Additionally, the instructional hypothesis can be complemented by the fact that greater awareness of inequality may experience feelings of injustice and relative deprivation since individuals with the same educational attainment have access to unequal rewards [@Adams1965; @Smith-eta2020_relative-deprivation]. In this sense, it has been shown that higher educational attainment is associated with higher perceived inequality [@Castillo2012a_justice]. Thus, this evidence leads us to suggest the following:

> $H_{\text{2b}}$: education is negatively associated with perceived meritocracy.

## Occupational Social Class

The relationship between structural position and meritocracy has mainly focused on income and educational attainment, setting occupational social class aside from the discussion. Occupational class not only represent the synthesis of educational achievement and income levels [@Ganzeboom1992], but it also represents a symbol of power and authority in contemporary societies [@Chan2007]. Social class schemes draw on occupational characteristics, autonomy and authority in the workspace as meaningful distinctions to categorise individuals [@Erikson1992], making them relatively homogeneous groups [@Connelly2016]. Social class also comprise individuals' experience in the labour market regarding job opportunities and wages [@Svallfors2006]. For example, lower classes are exposed to higher labour market risk, such as the likelihood of job loss and lower wages, leading individuals to experience deprivation due to their risk exposure [@rehm_risks_2009].

The literature has paid little attention to occupational social class on meritocracy perception. For example, @Smith2012 found that individuals belonging to the petty bourgeoisie perceive higher levels of meritocracy than wage earners. These findings are supported by @Sandoval2017, which found that skilled and service workers attribute more importance to effort in achieving success than unskilled workers. However, @Canales2015lse found that upwardly mobile individuals in the service class perceived lower levels of meritocracy.

Even when individuals have improved their social class position from lower to non-manual or service classes [@Torche2005a], the stagnation of wages and persistent economic inequality [@EspinozaNunez2014] can lead to negatively perceived meritocracy due to frustrated economic prospects [@barozet_clase_2011] or by experiences of discrimination in the labour market [@cea_discriminacion_2020]. Achieving higher occupational social class positions may not always result in satisfactory outcomes, which can further negative meritocracy perceptions when confronted with labour market constraints [@Canales2015lse]. In this regard, we argue that occupational social class position contributes with information about the labour market, such as job opportunities and wages, leading us to suggest that:

> $H_{\text{2c}}$: Working in non-manual and service occupations is negatively associated with perceived meritocracy.

## Interplay of subjective and objective reality

@@

Higher-income individuals tend to have higher inconsistencies in perceived status [@Castillo2013]. For example, income, education, and occupational prestige have a positive non-linear relationship with subjective social status, with higher-status individuals showing more variation in subjective status [@Andersson2018a]. In addition, income level and educational attainment inconsistencies have been linked to political preferences [@Kreckel1993, @Lenski1967], and discrepancies between class identity and social class may also influence political preferences [@Sosnaud2013], but socio-demographic controls do not consistently support these results.

Studies on subjective social status as a moderating factor have primarily focused on its relationship to other subjective variables rather than how it interacts with income or education. One study using data from the International Social Survey Programme (ISSP) found that higher perceptions of inequality lead to higher preferences for redistribution and lower perceptions of meritocracy and that the relationship between inequality and meritocracy perception becomes weaker for those who perceive themselves as having a higher societal position [@Fatke2018]. Another study in European countries showed that economic inequality does not influence preferences for redistribution when controlled by perceived inequality but is positively moderated by subjective status [@Bobzien2019]. A study on bachelor students with high subjective social status found that those who perceive themselves as having a higher status and greater meritocracy tend to justify higher levels of economic inequality [@bernardo_meritocracy_2021].

@@

Studies have demonstrated that three factors influence perceptions of social status: socioeconomic status, economic and social capital, and moral perceptions of individual achievement [@macclure_naming_2020]. Individuals who identify as lower status tend to have low incomes, less education, and live in disadvantaged areas. However, some individuals who belong to middle or high socioeconomic strata may still identify as lower status due to living in affluent neighbourhoods or having higher educational credentials, which symbolise higher social and cultural capital [@macclure_naming_2020]. Those who identify as high status tend to consistently have higher socioeconomic status, with educational meritocracy often serving as a criterion for their social position [@macclure_naming_2020].

Subjective social status can reflect an individual's position in the social structure and their experience of economic inequality through comparisons with nearby reference groups. Therefore, we predict that subjective status differences can lead to perceived meritocracy differences within socioeconomic groups. Thus, we predict that:

@@

The data used correspond to the third wave of the Longitudinal Social Study of Chile [@coes_estudio_2018], a survey developed to analyse the evolution of conflict and social cohesion in Chilean society over time. The sampling is probabilistic, stratified, clustered and multistage, with 3748 participants aged between 18 and 75. An analytical sample of 2652 cases with complete information is used for the analyses.

@@

@@

The objective position is measured through three variables. First, _Household Income Decile_ is calculated using monthly per capita household income and categorised into deciles. Second, _Educational level_ corresponds to the five categories for educational attainment following the International Standard Classification of Education [@unesco_international_2012]. Finally, we use an adapted five-category version of the EGP class scheme [@Connelly2016], a commonly used measurement in social stratification research [@solis_class_2019]. The Stata package `isko` was used to compute the class scheme [@hendrickx_isko_2002].

@@

  df_desc<- elsoc_w03 %>%
    dplyr::select(merit,meffort,mtalent,
           ess,inc10h.imp,edcine2,egpisko05,lngap_perc,
           sexo,edad,ppolcat) %>% 
    dplyr::select(-meffort,-mtalent,
                  # -sexo,-edad,
                  # -ppolcat,-lngap_perc
                  )
  varlabs <- sjlabelled::get_label(df_desc) #recuperar labels
  df_desc <- na.omit(df_desc)
  set_label(df_desc) <- varlabs  
 st_options(lang = "en",
            footnote = paste0("Source: Authors calculation based on ELSOC 2018 (N = ",
            dim(df_desc)[[1]],")"),
            bootstrap.css = F,
            # custom.css = "input/css/dfsummary.css"
            )
 df<- dfSummary(df_desc,
                plain.ascii = FALSE,
                style = "grid",
                tmp.img.dir = "/tmp",
                graph.magnif = TRUE,
                headings = F,  # encabezado
                varnumbers = F, # num variable
                labels.col = T, # etiquetas
                na.col = F,    # missing
                graph.col = T, # plot
                valid.col = T,max.string.width = 50,
                max.distinct.values = 8 # n valido 
                # col.widths = c(1000,10,10,10)
                )
 df$Variable <- NULL # delete variable column
 df$Valid <- NULL
 summarytools::view(df,file = "output/tables/desc01.html") # Ver tabla en un archivo HTML
 webshot(url ="output/tables/desc01.html",
         file ="output/tables/desc01.png",
         vwidth = 700,vheight = 600) # Remover archivo HTML
```

@@

Considering the dependent variable as a continuous, stepwise ordinary least square regression models are used (OLS). First, models are estimated for each of the independent variables. Second, a series of estimated models consider the interaction between objective position and subjective status. Finally, interaction effects are presented using conditional linear predicted values <!-- and linear predictions -->.

@@

Figure \@ref(fig:corr-mat) shows a positive correlation between perceived meritocracy and subjective status, with a slightly stronger association for the indicator for effort (r = 0.08) compared to intelligence (r = 0.06). Perceived meritocracy has a weak positive correlation with income decile, but weak negative correlations with education and social class.

@@

# Dicussion
